\begin{acknowledgements}

Far more people have had a contributing impact on this dissertation than I could possibly acknowledge in a few pages. The research that follows is as much the culmination of the oblique but persistent influence of the supporting cast of characters as it is the culmination of the core protagonists' focused efforts. Despite the incomplete list to follow, I must express my deepest gratitude for the village that has shaped me as a person and as a thinker over the years during which this work was completed. 

\bigskip

I want to thank my family and Todd for their patience and support throughout the duration of this excursion. Everyone should be as lucky as I am to have people in her life who empower her to forge her own path, and to be different if different is what she needs to be. 

\bigskip
 
I thank my advisor, Yoon Lee, for her both exceptional and patient academic guidance. Her ability to balance succinct, economical expression of thought with acute attention to detail holds my true admiration. I owe much of the clarity in the exposition of this work to her. I also want to extend my gratitude to the other members of my committee, Kate Calder and Sebastian Kurtek for their time and thoughtful feedback. I owe much of the conclusion of this work to Kate, whose transparent and honest advice provided perspective at a time when I most needed it. 

\bigskip

I have learned as much from my fellow graduate students as anyone else at Ohio State. I feel incredibly lucky to have been a part of a program in which the sense of competition between the students is replaced with a sense of community and teamwork. There is a warm, fuzzy spot in my heart for you guys and the slurry of Cockins Hall memories, wonderful and horrible and everything in between. 

\bigskip

My most sincere appreciation goes to Chris Holloman and my team at ICC. It is a rare opportunity to work with a group of individuals who make work as much of an opportunity to have fun it is an opportunity for growth. They have made simultaneously working while finishing this paper as easy as it could possibly be. You guys are the best.

\bigskip

This research is supported in part by the National Science Foundation under grants DMS-15-13566 and DMS-16-13110. 

\end{acknowledgements}

